---
title: "Introduction to boostPM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to boostPM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  fig.align = "center"
)
```

The `boostPM` package estimates probability distributions with an ensemble
of tree-based probability measures. This vignette fits a two-dimensional
distribution, evaluates the fitted density on a grid, and compares observations
generated from the fit with the training data.

## Data and support

Observations occupy rows and variables occupy columns. The example uses a
Gaussian copula to create dependent variables with beta marginal
distributions. The support is the unit square. Training observations must be
strictly inside a user-supplied support matrix.

```{r data}
library(boostPM)

set.seed(1)
n_training <- 200L
dependence <- 0.7
latent_1 <- rnorm(n_training)
latent_2 <- dependence * latent_1 +
  sqrt(1 - dependence^2) * rnorm(n_training)

training_data <- cbind(
  x1 = qbeta(pnorm(latent_1), shape1 = 2, shape2 = 5),
  x2 = qbeta(pnorm(latent_2), shape1 = 5, shape2 = 2)
)
support <- cbind(c(0, 0), c(1, 1))

plot(
  training_data,
  xlim = support[1, ],
  ylim = support[2, ],
  asp = 1,
  pch = 16,
  col = grDevices::adjustcolor("black", alpha.f = 0.45),
  xlab = "x1",
  ylab = "x2",
  main = "Training data"
)
```

## Fit an ensemble

The number of marginal and dependence trees controls the two fitting stages.
The settings below are deliberately moderate so that the vignette remains
suitable for routine package checks. Applied analyses should select tuning
parameters using held-out predictive performance.

```{r fit}
set.seed(2)
fit <- fit_boostpm(
  training_data,
  add_noise = FALSE,
  Omega = support,
  ntree_max_marginal = 6,
  ntree_max_dependence = 12,
  c0 = 0.1,
  gamma = 0.5,
  max_resol = 2,
  min_obs = 5,
  nbins = 6
)
```

Calling `set.seed()` controls the stochastic tree-fitting decisions. The same
seed and runtime provide the intended reproducibility level for a fit.

## Inspect the fitted object

`fit_boostpm()` is silent by default. Set `progress = "stage"` when
stage-level progress messages are useful. The S3 methods provide a compact
summary and diagnostic plots after fitting.

```{r inspect}
print(fit)
summary(fit)
plot(fit, type = "variable_importance")
```

Tree-size and maximum-depth diagnostics are also available.

```{r plot-tree-size}
plot(fit, type = "tree_size")
```

## Evaluate and display the fitted density

`predict()` returns log density by default. Here the fitted density is
evaluated over a regular grid covering the support. The heat map and contours
show the estimated joint density, with the training observations overlaid.

```{r evaluate-density}
grid_size <- 40L
x1_grid <- seq(0, 1, length.out = grid_size)
x2_grid <- seq(0, 1, length.out = grid_size)
evaluation_grid <- expand.grid(x1 = x1_grid, x2 = x2_grid)

fitted_density <- predict(
  fit,
  as.matrix(evaluation_grid),
  type = "density"
)
density_surface <- matrix(
  fitted_density,
  nrow = length(x1_grid),
  ncol = length(x2_grid)
)

image(
  x1_grid,
  x2_grid,
  density_surface,
  col = grDevices::hcl.colors(30, palette = "YlOrRd", rev = TRUE),
  xlab = "x1",
  ylab = "x2",
  main = "Fitted joint density",
  asp = 1
)
contour(
  x1_grid,
  x2_grid,
  density_surface,
  add = TRUE,
  drawlabels = FALSE,
  col = "grey30"
)
points(
  training_data,
  pch = 16,
  cex = 0.45,
  col = grDevices::adjustcolor("black", alpha.f = 0.45)
)
```

Points outside the support have log density `-Inf` and density zero.

```{r outside-support}
outside <- rbind(c(-0.10, 0.50), c(0.50, 0.50), c(1.10, 0.50))
predict(fit, outside, type = "log_density")
```

## Generate and compare samples

The fitted tree ensemble is also a generative model. The following code draws
1,000 observations and compares their joint distribution with the training
data using identical axes.

```{r simulate, fig.width=7, fig.height=3.6}
set.seed(3)
generated_data <- simulate(fit, nsim = 1000)

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(
  training_data,
  xlim = support[1, ],
  ylim = support[2, ],
  asp = 1,
  pch = 16,
  cex = 0.7,
  col = grDevices::adjustcolor("black", alpha.f = 0.45),
  xlab = "x1",
  ylab = "x2",
  main = "Training data"
)
plot(
  generated_data,
  xlim = support[1, ],
  ylim = support[2, ],
  asp = 1,
  pch = 16,
  cex = 0.45,
  col = grDevices::adjustcolor("#2166AC", alpha.f = 0.35),
  xlab = "x1",
  ylab = "x2",
  main = "Generated data"
)
par(old_par)
```

The marginal empirical distribution functions provide a second comparison.
Close curves indicate that the generated sample preserves the fitted marginal
behavior of the training data.

```{r compare-marginals, fig.width=7, fig.height=3.6}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
for (j in seq_len(ncol(training_data))) {
  plot(
    stats::ecdf(training_data[, j]),
    do.points = FALSE,
    verticals = TRUE,
    xlim = support[j, ],
    lwd = 2,
    col = "black",
    xlab = colnames(training_data)[j],
    ylab = "Empirical cumulative probability",
    main = paste("Marginal", colnames(training_data)[j])
  )
  lines(
    stats::ecdf(generated_data[, j]),
    verticals = TRUE,
    lwd = 2,
    col = "#2166AC"
  )
  legend(
    "bottomright",
    legend = c("Training", "Generated"),
    col = c("black", "#2166AC"),
    lwd = 2,
    bty = "n"
  )
}
par(old_par)
```

These plots are descriptive checks rather than a formal goodness-of-fit test.
Held-out log density or other problem-specific validation remains useful for
applied work.

## Practical notes

- Use a finite numeric matrix. Constant columns are not supported.
- `Omega` must have two columns and positive width in each dimension.
- `nbins` controls the uniform candidate split grid and must be at least two.
- `c0` is strictly between zero and one; `gamma` is non-negative.
- Larger experiments and tuning should use held-out predictive log density,
  following the numerical evaluation in Awaya and Ma (2024).

## References

Awaya, N. and Ma, L. (2024). Unsupervised Tree Boosting for Learning
Probability Distributions. *Journal of Machine Learning Research*, 25, 1--52.
