---
title: "Introduction to boostPM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to boostPM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>")
```

The `boostPM` package estimates probability distributions with an ensemble of
tree-based probability measures. This vignette fits a small model, evaluates
its density, and generates samples from the fitted distribution.

## Data and support

Observations occupy rows and variables occupy columns. This example draws two
variables from beta distributions with different shapes. The support is the
unit square. Training observations must be strictly inside a user-supplied
support matrix.

```{r data}
library(boostPM)

set.seed(1)
data <- cbind(
  rbeta(80, shape1 = 2, shape2 = 5),
  rbeta(80, shape1 = 5, shape2 = 2)
)

support <- cbind(c(0, 0), c(1, 1))
```

## Fit a small ensemble

The number of marginal and dependence trees controls the two fitting stages.
This deliberately small fit is suitable for an example, rather than a tuning
recommendation for an applied analysis.

```{r fit}
set.seed(2)
fit <- fit_boostpm(
  data,
  add_noise = FALSE,
  Omega = support,
  ntree_max_marginal = 2,
  ntree_max_dependence = 2,
  c0 = 0.1,
  gamma = 0.5,
  max_resol = 1,
  min_obs = 2,
  nbins = 4
)

fit$tree_size_store
fit$variable_importance
```

Calling `set.seed()` controls the stochastic tree-fitting decisions. The same
seed and runtime provide the intended reproducibility level for a fit.

## Inspect the fitted object

`fit_boostpm()` returns silently. The S3 methods provide a compact summary
and diagnostic plots.

```{r inspect}
print(fit)
summary(fit)
plot(fit, type = "variable_importance")
```

Tree-size and maximum-depth diagnostics are also available.

```{r plot-tree-size}
plot(fit, type = "tree_size")
```

## Evaluate density

`predict()` returns log density by default. Set `type = "density"` for density
values, or `type = "details"` for the cumulative mean log-density path.

```{r evaluate}
evaluation_points <- rbind(
  c(0.25, 0.75),
  c(0.50, 0.50),
  c(0.75, 0.25)
)

log_density <- predict(fit, evaluation_points, type = "log_density")
cbind(evaluation_points, density = exp(log_density))
```

Points outside the support have log density `-Inf`.

```{r outside-support}
outside <- rbind(c(-0.10, 0.50), c(0.50, 0.50), c(1.10, 0.50))
predict(fit, outside, type = "log_density")
```

## Generate samples

The fitted tree ensemble is also a generative model. Sampling uses R's
random-number generator, so it can be reproduced with a seed.

```{r simulate}
set.seed(3)
simulate(fit, nsim = 6)
```

## Practical notes

- Use a finite numeric matrix. Constant columns are not supported.
- `Omega` must have two columns and positive width in each dimension.
- `nbins` controls the uniform candidate split grid and must be at least two.
- `c0` is strictly between zero and one; `gamma` is non-negative.
- Larger experiments and tuning should use held-out predictive log density,
  following the numerical evaluation in Awaya and Ma (2024).

## References

Awaya, N. and Ma, L. (2024). Unsupervised Tree Boosting for Learning
Probability Distributions. *Journal of Machine Learning Research*, 25, 1--52.
